---
title: "ESM 244 Lab 1"
author: "Patrick Pelegri-O'Day"
date: "1/7/2022"
output: html_document
---

```{r setup, include=TRUE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F)

library(tidyverse)
library(here)
library(sf)
library(tmap)

### update.packages(ask = F)
```

### Read in the data

```{r}
sf_trees <- read_csv(here('data', 'sf_trees', 'sf_trees.csv'),
                     show_col_types = F)
```

### Part 1: wrangling and ggplot review

**Example 1:** Find counts of observations by `legal_status` & wrangle a bit.

```{r}
### method 1: group_by() %>% summarize()
sf_trees %>% 
  group_by(legal_status) %>% 
  summarize(tree_count = n())

### method 2:different way plus a few new functions
top_5_status <- sf_trees %>% 
  count(legal_status) %>% 
  drop_na(legal_status) %>% 
  rename(tree_count = n) %>% 
  relocate(tree_count) %>% # bring the column in parentheses to become 1st column
  slice_max(tree_count, n = 5) %>% # select tree_count with five highest values
  arrange(-tree_count) # arrange in descending order (default is ascending, note minus sign)
# alternative: arrange(desc(tree_count)) - does same thing as line above
```

Make a graph of the top 5 from above:

```{r}
ggplot(data = top_5_status, aes(x = fct_reorder(legal_status, -tree_count), y = tree_count)) + # rct_reorder is reordering legal_status by negative tree_count
  geom_col(fill = 'darkgreen') +
  labs(x = "Legal status", y = "Tree count") +
  coord_flip() # flip columns so they're extending horizontally instead of vertically +
theme_classic()
```

**Example 2:** Only going to keep observations where legal status is "Permitted Site" and caretaker is "MTA", and store as `permitted_data_df`

```{r}
# sf_trees$legal_status %>% unique()
# unique(sf_trees$caretaker) - same effect as above

permitted_data_df <- sf_trees %>% 
  filter(legal_status == 'Permitted Site', caretaker == 'MTA')
```

**Example 3:** Only keep Blackwood Acacia trees, then only keep columns `legal_status`, `date`, `latitute`, `longitude`, and store as `blackwood_acacia_df`

```{r}
blackwood_acacia_df <- sf_trees %>% 
  filter(str_detect(species, 'Blackwood Acacia')) %>%  # str_detect finds observations in column species where any part of the character value has 'Blackwood Acacia' in it; there can be other character values in there too
  select(legal_status, date, lat = latitude, lon = longitude)

### Make a graph of locations
ggplot(blackwood_acacia_df, aes(x = lon, y = lat))
```




